Replacing standard galaxy profiles with \\ mixtures of Gaussians

### 2012-09-19 documents/mixture_models.tex
intro sort-of done

\newcommand{\foreign}[1]{\textit{#1}}
\newcommand{\etal}{\foreign{et\,al.}}
\newcommand{\documentname}{\textsl{Note}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\tractor}{\project{Tractor}}
\newcommand{\sdss}{\project{SDSS}}

@@

We need good performance in two-dimensional image fitting because,
with \thetractor (Lang \etal, forthcoming), we are building a
comprehensive model of all the imaging data we can; this will have
models of many millions of galaxies in imaging that contains on the
order of $10^{13}$ pixels.  We are basing our models on the
\sdss\ models (\citealt{lupton})...Limit scope of this document to
just what we need for the \tractor...

...somehow have to cite GALFIT (\citealt{galfit}) in this section...

...We need to put somewhere Lupton-like comments about sampling the
pixel-convolved-PSF-convolved scene...

In what follows we will make heavy use of two-dimensional Gaussian or
Normal distributions, which look like
\begin{eqnarray}\displaystyle
\normal(\pos\given\mean,\var) &\equiv& \frac{1}{2\pi}\,\det(\var)^{-1/2}\,\exp(-\frac{1}{2}\,\transpose{[\pos-\mean]}\cdot\inverse{\var}\cdot[\pos-\mean])
\quad ,
\end{eqnarray}
where $\pos$ and $\mean$ are two-dimensional vectors (usually in the
focal plane or on the sky or something like that), and $\var$ is a
symmetric $2\times 2$ variance tensor or matrix, and implicitly the
vectors are column vectors.  A \emph{mixture of Gaussians} is a linear
superposition of Gaussians.  Any positive two-dimensional function
with finite support and finite total integral---including as a special
case any two-dimensional probability distribution function---can be
represented as a mixture of Gaussians to arbitrary accuracy; that is
\begin{eqnarray}
p(\pos) &\approx& \sum_{k=1}^K a_k\,\normal(\pos\given\mean_k,\var_k)
\\
1 &=& \sum_{k=1}^K a_k
\quad ,
\end{eqnarray}
where $p(\pos)$ is any probability distribution of a two-dimensional
quantity $\pos$, the $\approx$ symbol implies approximation, $K$ is
the number of Gaussians used in the mixture, and the $K$ Gaussians
have amplitudes $a_k$, means $\mean_k$, and variance tensors $\var_k$.
The sum-to-one condition ensures that the probability distribution
approximation is properly normalized.

We wish to make an approximation $Q_{\exp}(\cdot)$ to the
two-dimensional circular exponential profile
\begin{eqnarray}\displaystyle
Q_{\exp}(\spos) &\approx& \frac{1}{Z_{\exp}}\,\exp(-\frac{|\spos|}{r_{\exp}})
\\
Z_{\exp} &=& X.XXXX
\\
r_{\exp} &=& X.XXXX
\quad ,
\end{eqnarray}
where $\spos$ is a dimensionless focal-plane position, and $Z_{\exp}$
and $r_{\exp}$ are parameters that ensure that the profile integrates
to unit flux and has unit half-light radius.  The position $\spos$ is
dimensionless because it parameterizes the unit-size and unit-flux
dimensionless function.  We seek the best (where ``best'' will be
defined below) $M_g$-Gaussian mixture approximation
\begin{eqnarray}\displaystyle
Q_{\exp}(\spos) &=& \sum_{m=1}^{M_g} g_m\,\normal(\spos\given\zero,\Gm_m)
\\
1 &=& \sum_{m=1}^{M_g} g_m
\\
\Gm_m &=& \sigma^2_{\exp,m}\,\identity
\quad ,
\end{eqnarray}
where all of the means are exactly zero and all of the variances
$\Gm_m$ in the mixture are scalar multiples of the identity matrix
$\identity$ because we are requiring this dimensionless function to be
precisely circular (so every component is itself circular and
concentric).  Similarly for the de~Vaucouleurs profile
\begin{eqnarray}\displaystyle
Q_{\dev}(\spos) &\approx& \frac{1}{Z_{\dev}}\,\exp(-\left[\frac{|\spos|}{r_{\dev}}\right]^{1/4})
\\
Z_{\dev} &=& X.XXXX
\\
r_{\dev} &=& X.XXXX
\\
Q_{\dev}(\spos) &=& \sum_{m=1}^{M_h} h_m\,\normal(\spos\given\zero,\Hm_m)
\\
1 &=& \sum_{m=1}^{M_h} h_m
\\
\Hm_m &=& \sigma^2_{\dev,m}\,\identity
\quad .
\end{eqnarray}

...Cite (\citealt{ciotti}) re half-light radius constants...

...Lupton profile as different from de~Vaucouleurs...Sersic profiles...

...Discuss the cutoffs and regularization used by PHOTO and adjust $Z$ and $r$ accordingly...

>>>>

\newcommand{\foreign}[1]{\textit{#1}}
\newcommand{\etal}{\foreign{et\,al.}}
\newcommand{\documentname}{\textsl{Note}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\thetractor}{\project{The~Tractor}}
\newcommand{\sdss}{\project{SDSS}}

@@

Along those lines, it is important in any image-modeling situation to
think of the PSF as the \emph{pixel-convolved} point-spread function.
Under this choice, synthesis of a pixelized image involves only
convolution with the PSF and sampling at the pixel centers.  That is,
image synthesis (returned to at the end of this \documentname)
consists of PSF convolution (an arithmetic operation on the two
mixtures of Gaussians) followed by sampling (evaluation of the
resulting mixture at the pixel centers).

We need good performance in two-dimensional image synthesis (for
fitting) because, with \thetractor\ (Lang \etal, forthcoming), we are
building a comprehensive model of all the imaging data we can; this
will have models of many millions of galaxies in imaging that contains
on the order of $10^{13}$ pixels.  We are basing our models on the
\sdss\ Catalog galaxy models, which only include exponential,
de~Vaucouleurs, and composite (mixture of the two) radial profiles.
In detail, in fact, the \sdss\ Catalog models are small modifications
of these (\citealt{lupton}); details below.  So our goal here is to
provide replacements for these models in order to improve the
performance of galaxy image modeling and analysis software.  Some of
the models we present have been used previously in the study of XXX
(\cite{bundy}); they could also be used to speed other image-fitting
systems, like the very successful \project{GALFIT} (\citealt{galfit}).

Because we are thinking about two-dimensional imaging, we use here
two-dimensional Gaussian or Normal distributions, which look like
\begin{eqnarray}\displaystyle
\normal(\pos\given\mean,\var) &\equiv& \frac{1}{2\pi}\,\det(\var)^{-1/2}\,\exp(-\frac{1}{2}\,\transpose{[\pos-\mean]}\cdot\inverse{\var}\cdot[\pos-\mean])
\quad ,
\end{eqnarray}
where $\pos$ and $\mean$ are two-dimensional vectors (usually in the
focal plane or on the sky or something like that), and $\var$ is a
symmetric $2\times 2$ variance tensor or matrix, and implicitly the
vectors are column vectors.  A \emph{mixture of Gaussians} is a linear
superposition of Gaussians.  Any positive two-dimensional function
with finite support and finite total integral---including as a special
case any two-dimensional probability distribution function---can be
represented as a mixture of Gaussians to arbitrary accuracy; that is
\begin{eqnarray}
p(\pos) &\approx& \sum_{k=1}^K a_k\,\normal(\pos\given\mean_k,\var_k)
\\
1 &=& \sum_{k=1}^K a_k
\quad ,
\end{eqnarray}
where $p(\pos)$ is any probability distribution of a two-dimensional
quantity $\pos$, the $\approx$ symbol implies approximation, $K$ is
the number of Gaussians used in the mixture, and the $K$ Gaussians
have amplitudes $a_k$, means $\mean_k$, and variance tensors $\var_k$.
The sum-to-one condition ensures that the probability distribution
approximation is properly normalized.

We wish to make an approximation $Q_{\exp}(\cdot)$ to the
two-dimensional circular exponential profile
\begin{eqnarray}\displaystyle
Q_{\exp}(\spos) &\approx& \frac{1}{Z_{\exp}}\,\exp(-\frac{|\spos|}{\xi_{\exp}})
\\
Z_{\exp} &=& X.XXXX
\\
\xi_{\exp} &=& X.XXXX
\quad ,
\end{eqnarray}
where $\spos$ is a dimensionless focal-plane position, and $Z_{\exp}$
and $\xi_{\exp}$ are parameters that ensure that the profile integrates
to unit flux and has unit half-light radius.  The position $\spos$ is
dimensionless because it parameterizes the unit-size and unit-flux
dimensionless function.  We seek the best (where ``best'' will be
defined below) $M_g$-Gaussian mixture approximation
\begin{eqnarray}\displaystyle
Q_{\exp}(\spos) &=& \sum_{m=1}^{M_g} g_m\,\normal(\spos\given\zero,\Gm_m)
\\
1 &=& \sum_{m=1}^{M_g} g_m
\\
\Gm_m &=& \sigma^2_{\exp,m}\,\identity
\quad ,
\end{eqnarray}
where all of the means are exactly zero and all of the variances
$\Gm_m$ in the mixture are scalar multiples of the identity matrix
$\identity$ because we are requiring this dimensionless function to be
precisely circular (so every component is itself circular and
concentric).  Similarly for the de~Vaucouleurs profile
\begin{eqnarray}\displaystyle
Q_{\dev}(\spos) &\approx& \frac{1}{Z_{\dev}}\,\exp(-\left[\frac{|\spos|}{\xi_{\dev}}\right]^{1/4})
\\
Z_{\dev} &=& X.XXXX
\\
\xi_{\dev} &=& X.XXXX
\\
Q_{\dev}(\spos) &=& \sum_{m=1}^{M_h} h_m\,\normal(\spos\given\zero,\Hm_m)
\\
1 &=& \sum_{m=1}^{M_h} h_m
\\
\Hm_m &=& \sigma^2_{\dev,m}\,\identity
\quad .
\end{eqnarray}
...Cite (\citealt{ciotti}) re half-light radius constants...

The Lupton profiles are...

...Lupton profiles as different from de~Vaucouleurs...Sersic profiles...Discuss the cutoffs and regularization used by PHOTO and adjust $Z$ and $r$ accordingly...
